Human capital mediates natural selection in contemporary humans

### 2022-01-19 why-natural-selection.Rmd
Add Age/Qualification weights to all regressions in this branch

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

library(drake)
library(magrittr)
library(dplyr)
library(forcats)
library(ggplot2)
library(tidyr)
library(santoku) # keep this after tidyr as it masks chop
library(purrr)
library(glue)
library(broom)
library(huxtable)
loadNamespace("scales")

@@

knitr::opts_chunk$set(
        echo = FALSE, 
        warning = FALSE, 
        cache = FALSE, 
        error = TRUE,
        fig.height = 3.5
      )

@@

my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "#FAC32D")

@@

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)

@@

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
          !! score_col := pretty_names(!! score_col),
          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
        ) %>% 
        rename(p = p.value) %>% 
        ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
          my_vline + 
          confint_geom_segment +
          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
          pval_scale() +
          my_fill_scale(n = n_cats, direction = fill_direction) +
          my_labs() +
          my_theme() +
          scale_linetype_manual(name = "", values = c("95% c.i." = 1))

}

@@

pval_geom_point <- function (size = 1.5, stroke = 0.5, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i."), 
                    color = "grey60", alpha = 0.25, size = 1.1, 
                    lineend = "round")
}

@@

  fill_colors <- c("steelblue4", "#D63C3C")

  if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

@@

n_in_regs <- function (var, dv = "n_children", data = famhist) {
  fh_subset <- data
  if (dv == "n_children") fh_subset %<>% filter(kids_ss)
  # all PGS have the same NA pattern, so we just use one
  tbl <- table(fh_subset[
                  ! is.na(fh_subset[dv]) & ! is.na(fh_subset$whr_combined), 
                  var
                ])

@@

add_n <- function (var, famhist_var = var, dv = "n_children", data = famhist) {
  var <- as.factor(var)
  n <- n_in_regs(famhist_var, dv, data)
  levels(var) <- sprintf("%s (N = %s)", levels(var), 
                         prettyNum(n[seq_len(nlevels(var))], big.mark = ","))
  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
  # var_n[is.na(var)] <- NA
  var
}

@@

```{r NOTES-TODO, eval=FALSE}
"

# TODO response to PNAS

* Only use education, not first job pay, in "controlled" regressions; check
  description is correct - done.
* Add CIs - done.
* Clarify collider bias/endogeneity - done.
* Formalize theory - done.
  - Comment on results where effects are actually reversed. This works
    nicely if you throw in a money cost - done computational example.
* Clarify causality? - done...
  - Helps to deal with the "collider bias" argument of reviewer
  - Might also be a better test of mediation via human capital
  - One other possibility. If we use only people with first profession
    teacher, that (as it happens) shuts down the EA3-income correlation.
    It also shuts down the EA3-fertility relationship, suggesting that
    income is indeed the mediator.
* Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
  - Point is that we can explain how natural selection is occurring.
* "Discuss empirical predictions for AFLB grounded in dynamic theory"
  - hmm, tricky, maybe we can include it in discussion of economic results at
  end of formal section.
  - think our theory with 2 periods probably covers this
* Why don't women see stronger effects? But note, they do: not significant
  after correction for multiple testing, but a 25% difference. 
  - I think we leave this for reviewers.
* Quadratic regressions for parents - DONE

# Open questions/TODO

* Limit max no. of children in regressions (use an "in_children" bool)
  - not necessary. Very few have abnormally many.
* Consider doing separate regressions for "any children" vs. "n children | 1 or more"
  - heckit too slow and gives crazy answers, just use plm/ols
  - some variables do look different
* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Reweight other regressions (e.g. the omnibus regression in the appendix)?

* If you take "year of last job started" and see if it is after/before
  the first child, you can figure out whether mothers keep working.
  - better still, look at all jobs started and see if any are within e.g. 5 years of
  first child's birth
  - this could then be another way to get at income vs substitution effects. You'd
  predict negative selection effects among mothers who worked. Of course the
  choice to work is endogenous.
* double check which scores are reverse coded!

# WHAT MODEL FITS YOUR DATA
## ASSUMPTIONS
* Genetics produce human capital h. 
* h and education are compliments in producing earnings
* There are income and substitution effects on fertility:
  - income effects are stronger at high income levels (Becker and Tomes 1976)
    and among couples
* People seek partners when in the labour market. Women match a person, observe
  their quality, and decide whether to have a first child.
* Then they (may) have more children, either with the partner or not 
  (high quality partners stay around more)
* High quality partners increase household income

## RESULTS
* People with low h are more likely to leave education early and enter the labour market
  - because e and h are complements
  - this is part of the substitution effect, as they expect to spend some time
    raising children rather than being in work
* At low values of h, increases in h reduce fertility (substitution effect dominates)
  at higher values, smaller reductions or increases (income effect stronger)
  - because h predicts income
* In 1-parent households, increases in h reduce fertility (sub. effect), in 
  2-parent households, smaller reductions or increases (income effect stronger)
  - because h predicts income and because stronger income effect among couples
* People with low h start having children earlier (b/c leave educ earlier)
  - so part of the substitution effect comes via this
  - controlling for this the income effect dominates
  - hence increases in h reduce fertility less once we control for AFLB

## THOUGHTS
* Should we distinguish between men and women in couples?
  - within couples, male h should increase fertility (via earnings)
  - female h should decrease it (ditto?)
  - see explore-sex-vs-pay.R

## RIVAL MECHANISMS
* Lynn 2011 mentions two:
  - social class gradient on ability to delay gratification
  - modern society protects the weak (Darwin 1871, Spencer 1874)
    NB this latter doesn't predict more fertility among "dysgenic" ppl, just
    equal fertility and then genetic drift does its work
    - examples include modern sanitation, post-1800
    - and monogamy!
"
```

@@

Lastly, we discuss the effects of natural selection. While measured effects on
mean genetic scores are small, natural selection may substantially increase the
correlation between genetic scores and income, i.e. the "genetic lottery"
[@harden2021genetic].

@@

```{r calc-sex-ratio}
female_ratio <- mean(famhist$sex == 0)
```

@@

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green triangles show a significant linear increase over time (p < 0.05/33). Red squares show a significant decrease."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        fatten = 1.25) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c("-" = "square", "+" = "triangle", "o" = "circle open")) +
      theme(
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

To test more directly for natural selection, we regress respondents' number of
children ($N_i$) on their polygenic scores (PGS):

```{=tex}
\begin{equation}
N_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\label{eq:regression}
\end{equation}
```

The "selection effect", $\beta$, reflects the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization] Note that equation
\eqref{eq:regression} does not control for many environmental and genetic
factors that could affect fertility, and as a result, $\beta$ is not an estimate
of the causal effect of a polygenic score on fertility. However, natural
selection is a correlational phenomenon: polygenic scores which correlate with
high fertility are being selected for, whatever the underlying causal
relationship.

[^normalization]: The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
where $N$ is the number of children. Since PGS are normalized to variance 1 and
mean 0, this reduces to 
$Cov(N, PGS) \equiv E(N\ PGS) - E(N) E(PGS) = E(N\ PGS)$. This is the polygenic 
score weighted by the number of children, which is the average polygenic score 
in the next generation.

@@

res_wt_combined <- bind_rows(
  None                = res_all %>% 
                          filter(
                            reg.type == "controlled", 
                            dep.var == "N children"
                          ) %>% 
                          select(-reg.type, -dep.var, -score_name),
  Geographical        = res_wt_msoa_weights,
  `Age/Qualification` = res_wt_age_qual_weights,
  `Age/Qual/AFLB`     = res_wt_flb_weights,
  .id = "Weights"
)

@@

To correct for ascertainment bias, we weight participants using population data.
We try three alternative weighting schemes: (1) "Geographical" weighting, by
geography, age and presence/absence of a partner; (2) age and highest
educational qualification; and (3) for women only, age, highest qualification,
and age at first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots
selection effects among the entire sample, estimated with the three weighting
schemes, and also unweighted. Effects for `r res_sig_all` out of 33 polygenic
scores are significant at $p < 0.05/33$ under all four weighting schemes, with
`r res_sig_any` scores reaching significance under at least one scheme.
Weighting makes a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic
variables.[^weighting]

@@

We also check for balancing and diversifying selection by estimating
\@ref(eq:regression) with a quadratic term.[^balance-diverse-def] In particular,
we find diversifying selection for educational attainment polygenic scores: at
higher values of these scores, the negative effect on fertility is smaller
(Appendix Figure \@ref(fig:plot-purifying)).

[^balance-diverse-def]: Balancing selection selects for intermediate values, 
reducing the variance of the trait under selection. Diversifying selection
selects for extreme values and increases the variance.

```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of number of children on a polygenic score."}

@@

res_wt_combined %>% 
      mutate(
        Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                    "Age/Qual/AFLB", "None"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.5, stroke = 0.5) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "None"              = "grey15", 
        "Geographical"      = "dodgerblue4",
        "Age/Qualification" = "steelblue2",
        "Age/Qual/AFLB"      = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "None" = "square filled", 
        "Geographical" = "circle filled",
        "Age/Qualification" = "circle filled", 
        "Age/Qual/AFLB" = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
      my_theme()

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income, education and sex, and
family structure variables including age at first live birth, presence of a partner, and
lifetime number of sexual partners.
We emphasize that these categories are not necessarily exogenous to the polygenic
scores. For example, both in the data and in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related
polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
differences in equation \@ref(eq:regression) across subgroups constrain the set
of possible explanations for natural selection. As we show below, a simple
economic model predicts the differences we find.

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category.  These results could be
explained by age, if older respondents have lower income and are less educated,
and also show more natural selection on polygenic scores. However, when we rerun
the regressions, interacting the polygenic score with income category and also
with a quadratic in age, the interaction with income remains significant at
$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
if we interact the PGS with age of leaving full time education and a quadratic
in age, the interaction with age leaving full time education remains significant
at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.

Selection effects are also different between men and women (Appendix Figure
\@ref(fig:plot-sexes)). Differences are particularly large for educational
attainment, height and MDD. Several polygenic scores for mental illness and
personality traits are more selected for (or less against) among women,
including major depressive disorder (MDD), schizophrenia and neuroticism, while
extraversion is more selected for among men. Scores for waist circumference and
waist-hip ratio are less selected for among women. One
possible reason for these sex differences is that polygenic scores may affect
fertility via success in marriage markets, and men and women may value different
characteristics in these.

@@

We next focus on variables related to household type and reproductive
strategy. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
separately for each tercile of AFLB. Several effects are strikingly different
across terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several polygenic
scores for body measurements are selected against only among older mothers. The
correlation between effect sizes for the youngest and oldest terciles is 
`r cor_age_flb`.

[^aflb-no-men]: This information is unavailable for men.

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4}

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(-5, 0, 0, -85),
                  legend.margin        = margin(0, 0, 0, 0),
                  legend.box.just      = "left",
                  legend.justification = "left",
                  legend.spacing       = unit(1, "lines")
                )
sexes_guides <- guides(
                  # fill  = guide_legend(ncol = 1),
                  # color = guide_legend(ncol = 1)
                )

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = `N partners`, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
        sexes_guides

res_with_partner_sex %>% 
      mutate(
        score_name = pretty_names(score_name),
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = Household, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
        sexes_guides

@@

```{r}
drake::loadd(res_mediation)

>>>>

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "why-natural-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

library(drake)
library(magrittr)
library(dplyr)
library(forcats)
library(ggplot2)
library(tidyr)
library(santoku) # keep this after tidyr as it masks chop
library(purrr)
library(glue)
library(broom)
library(huxtable)
loadNamespace("scales")
loadNamespace("shades")

@@

knitr::opts_chunk$set(
        echo       = FALSE, 
        warning    = FALSE, 
        cache      = FALSE, 
        error      = FALSE,
        fig.height = 3.5
      )

@@

my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "#f3b372")

@@

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)
  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

@@

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/33)) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
    scale_shape_manual(values = 1:n_cats + 14) +
    my_labs() +
    my_theme() +
    scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
    scale_alpha_discrete(range = c(1, 0),
                         guide = guide_legend(
                           override.aes = list(
                             shape = c(1, 16),
                             color = "black",
                             alpha = 1,
                             size = 1.8
                           )
                         ))
}

@@

pval_geom_point <- function (size = 1.6, stroke = 0.8, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
                    color = "grey60", alpha = 0.25, size = 1.1, 
                    lineend = "round")
}

@@

  fill_colors <- c("steelblue4", "darkred")

  if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    shades::lightness(scale_color_brewer(aesthetics = aesthetics, ...), 
                        shades::scalefac(0.85))
  }
}

@@

n_in_regs <- function (var, dv = "RLRS", data = famhist) {
  fh_subset <- data
  if (dv == "RLRS") fh_subset %<>% filter(kids_ss)
  # all PGS have the same NA pattern, so we just use one
  tbl <- table(fh_subset[
                  ! is.na(fh_subset[dv]) & ! is.na(fh_subset$whr_combined), 
                  var
                ])

@@

add_n <- function (var, famhist_var = var, dv = "RLRS", data = famhist) {
  var <- as.factor(var)
  n <- n_in_regs(famhist_var, dv, data)
  levels(var) <- sprintf("%s (N = %s)", levels(var), 
                         prettyNum(n[seq_len(nlevels(var))], big.mark = ","))
  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
  # var_n[is.na(var)] <- NA
  var
}

@@

@@

Lastly, we discuss the effects of natural selection. While measured effects on
mean genetic scores are small, natural selection may substantially increase the
correlation between genetic scores and income, increasing genetic differences 
between different social groups, and thus making the "genetic lottery"
[@harden2021genetic] more unfair.

@@

@@

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Circles are means for 5-year intervals. Lines are 95% confidence intervals. Solid circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey60", 
                       size = 3) +
      geom_point(aes(y = score), size = 2) +
      facet_wrap(~score_name, ncol = 6) +
      my_hline +
      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "circle open")) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

To test more directly for natural selection, we regress respondents' relative 
lifetime reproductive success (RLRS) on their polygenic scores (PGS):

```{=tex}
\begin{equation}
\mathrm{RLRS}_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\label{eq:regression}
\end{equation}
```

RLRS is defined as respondents' number of children, divided by the mean number
of children of people born in the same year. The "selection effect", $\beta$,
reflects the strength of natural selection within the sample. In fact, since
polygenic scores are normalized, $\beta$ is the expected polygenic score among
children of the sample [@beauchamp2016genetic].[^normalization] Note that
equation \eqref{eq:regression} does not control for many environmental and genetic
factors that could affect fertility, and as a result, $\beta$ is not an estimate
of the causal effect of a polygenic score on fertility. However, natural
selection is a correlational phenomenon: polygenic scores which correlate with
high fertility are being selected for, whatever the underlying causal
relationship.

[^normalization]:The selection effect $\beta$ equals $Cov(RLRS, PGS)/Var(PGS)$.
Since PGS are normalized to variance 1 and mean 0, this reduces to $Cov(RLRS,
PGS) = E(RLRS\times PGS) - E(RLRS) E(PGS) = E(RLRS \times PGS)$. This is the
polygenic score weighted by relative lifetime reproductive success, which is the
average polygenic score in the next generation [@robertson1966mathematical].

@@

res_wt_combined <- bind_rows(
  None                = res_all %>% 
                          filter(
                            reg.type == "controlled", 
                            dep.var == "RLRS"
                          ) %>% 
                          select(-reg.type, -dep.var, -score_name),
  Geographical        = res_wt_msoa_weights,
  `Age/Qualification` = res_wt_age_qual_weights,
  `Age/Qual/AFLB`     = res_wt_flb_weights,
  .id = "Weights"
)

@@

To correct for ascertainment bias, we weight participants using population data.
We try three alternative weighting schemes: (1) "Geographical" weighting, by
geography, age and presence/absence of a partner; (2) age and highest
educational qualification; and (3) for women only, age, highest qualification,
and age at first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots
selection effects among the entire sample, estimated with the three weighting
schemes, and also unweighted. Effects for `r res_sig_all` out of 33 polygenic
scores are significant at $p < 0.05/33$ under all four weighting schemes, with
`r res_sig_any` scores reaching significance under at least one scheme.[^balance-diverse-def]
Weighting makes a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic
variables.[^weighting]

[^balance-diverse-def]: We also check for stabilizing and disruptive selection by
estimating \@ref(eq:regression) with a quadratic term. Stabilizing selection
selects for intermediate values, while disruptive selection selects for extreme 
values. In particular, we find disruptive selection for educational attainment
polygenic scores: at higher values of these scores, the negative effect on
fertility is smaller (Appendix Figure \@ref(fig:plot-purifying)).

@@

```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of number of children on a polygenic score. P value threshold is 0.05, Bonferroni-corrected for multiple comparisons. Confidence intervals are uncorrected."}

@@

res_wt_combined %>% 
      mutate(
        Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                    "Age/Qual/AFLB", "None"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.6, stroke = 0.8, alpha = 0.9) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "None"              = "grey15", 
        "Geographical"      = "dodgerblue4",
        "Age/Qualification" = "steelblue2",
        "Age/Qual/AFLB"      = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "None" = "square filled", 
        "Geographical" = "circle filled",
        "Age/Qualification" = "circle filled", 
        "Age/Qual/AFLB" = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
      my_theme()

@@

We now show the empirical puzzles which motivate our economic model. Each
concerns differences in the strength of natural selection across different
subgroups in the sample. Splitting the sample by demographic and social
variables, we re-estimate \@ref(eq:regression) within each subgroup. We use
socio-economic variables including income and education, and family structure
variables including age at first live birth, presence of a partner, and lifetime
number of sexual partners.

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category. These results are robust
to controlling for respondents' age (Appendix section \@ref(sec:age-control)).

@@

Turning to family structure, we split respondents by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
Selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).[^sex]

[^sex]: The same pattern holds if we analyse men and women separately (Appendix Figure
\@ref(fig:plot-n-partners-sex)). We also directly compared selection effects 
between men and women (Appendix Figure \@ref(fig:plot-sexes)).

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and presence of a partner.", fig.subcap = c("Lifetime number of sexual partners", "Presence of a partner"), fig.ncol = 1, fig.align = "center", fig.height = 4}

drake::loadd(res_partners_joint)
res_partners_joint %<>%       
      filter(grepl(":", term)) %>% 
      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))

n_regs <- as.double(nrow(res_partners_joint))
res_partners_joint %>% 
      standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
                        score_col = score_name)

drake::loadd(res_with_partner)
res_with_partner %<>% filter(grepl(":", term))

n_regs <- as.double(nrow(res_with_partner))

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      standard_ggplot(fill_col = Household, n_regs = n_regs, 
                        score_col = score_name)
```

@@

@@

```{r calc-mediation}
drake::loadd(res_mediation)
